|

### 2021-08-10 prova_knit_finale.Rmd
Aggiunta parte SV aggiornata

SC<-matrix(nrow=2,ncol=3)
rownames(SC)<-c("Mean","Standard Deviation")
colnames(SC) <- c("S&P500", "Dow Jones", "STOXX50E")

@@

We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis. Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows. For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$. Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \

First, a general point is that the sequence of percentage log returns are different across the three indices. The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents (SEE SECTION X). In particular, as previously noticed, there are extensive similarities between the DOW JONES Index and the S&P500, which can be visualized when looking at the percentage of log returns in the corresponding plots, since both are describing the US Economy. However, the STOXX50E differs quite extensively, especially in the different sensitivity to the Covid-19 Crisis and its consequences,  representing the European Economy. Aside from this note, the considerations expressed below on the sequences "filtered volatilities"\footnote{Per ora nei grafici c'è scritto "Filtered States", verrà corretto.} for the various techniques hold almost uniformly for all the three indices. Therefore, in the next paragraph we will be referring to the set of plots pertaining to one specific index (which could be any of the three).\

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza. Inoltre, va chiarito il punto della "doppia convergenza".} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles. Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below. Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}

```{r include=FALSE}
Errorcomp1<-matrix(NA,ncol=6,nrow=10)
colnames(Errorcomp1)<-c("N","BPF", "GPFOPT", "APF", "BAPF","SIS")
rownames(Errorcomp1)<-c("RMSE","RMSE","RMSE","RMSE","RMSE","MAE","MAE","MAE","MAE","MAE")
Errorcomp1[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
Errorcomp1[,2]<-c(0.4233094,0.15689352,0.06901099,0.03817009,0.03045272,0.2807802,0.09658088,0.03312132,0.01889245,0.01455814)
Errorcomp1[,3]<-c(0.3685259,0.15094281,0.07669050,0.03568058,0.02709059,0.2796141,0.09623692,0.03696833,0.01664805,0.01422964)
Errorcomp1[,4]<-c(0.5478362,0.2404408,0.08878446,0.05742351,0.04295610,0.4310335,0.1756985,0.05538968,0.03186140,0.02448565)
Errorcomp1[,5]<-c(0.5595929,0.2348842,0.10295053,0.06109947,0.05151115,0.4377886,0.1839363,0.06726034,0.03448388,0.02726281)
Errorcomp1[,6]<-c(0.9168639,1.1055065,0.6250516,0.6200359,0.5118927,0.7275327,0.9179852,0.4971839,0.4475265,0.406363)

@@

Using the APF as the benchmark does not deliver a different picture than the one we previously had. Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at 10, 100 and 1000 particles.\
Its performance ameliorates dramatically at N=1000, while at N=5000 the halving of its RMSE and MAE even allows the APF to overtake the Guided PF, though falling short of the Bootstrap PF. At N=10000 both the Basic PF and the APF achieve lower RMSEs than the Bootstrap and the Guided PF, even though the MAE criterion still selects the Bootstrap as the closest to the target distribution.\
Interestingly, this time the performance of the Bootstrap filter worsens when N is increased from 5000 to 10000 (both according to RMSEs and MAEs, whose disagreements, once again, are overall infrequent). Such evidence suggests that, even at very large (yet finite) values of N, some difference between the approximated filtering distributions still persists across our particle filters.\

```{r include=FALSE}
Errorcomp3<-matrix(NA,ncol=2,nrow=2)
colnames(Errorcomp3)<-c("RMSE", "MAE")
rownames(Errorcomp3)<-c("N=50000", "N=100000")
Errorcomp3[1,]<-c(0.04407704,0.01667898)
Errorcomp3[2,]<-c(0.03575915,0.01476220)
```

@@

Johannes, M., Evers, L. (2007). Monte Carlo Methods. Lecture Notes. University of Bristol, Department of Mathematics.\

@@

Petris, G., Petrone, S., Campagnali, P. (2010). Dynamic Linear Models with R. Springer.\

>>>>

SC<-matrix(nrow=2,ncol=3)
rownames(SC)<-c("Mean","Standard Deviation")
colnames(SC) <- c("S&P500","Dow Jones","STOXX50E")

@@

We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis. Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows. For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t/2}$. Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \

```{r echo=FALSE, fig.align='center', fig.cap="Observations and filtered volatility, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}

# Filtered Volatility Plots

library(openxlsx)
library(ggplot2)
library(ggpubr)

realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
y1<-realdatasetSPX[,2]
timeframe<-c(1:length(y1))

sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
dfsv1<-data.frame(timeframe,y1,sv_bpf1)
dfsv2<-data.frame(timeframe,y1,sv_gpfopt1)
dfsv3<-data.frame(timeframe,y1,sv_apf1)
dfsv4<-data.frame(timeframe,y1,sv_lw1)
dfsv5<-data.frame(timeframe,y1,sv_bapf1)
dfsv6<-data.frame(timeframe,y1,sv_sis1)

Filtplot1<-function(dataframe,title){
  mean=sv_bpf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot2<-function(dataframe,title){
  mean=sv_gpfopt1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot3<-function(dataframe,title){
  mean=sv_apf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot4<-function(dataframe,title){
  mean=sv_lw1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot5<-function(dataframe,title){
  mean=sv_bapf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot6<-function(dataframe,title){
  mean=sv_sis1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotobs1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
plotobs2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
plotobs3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
plotobs4<-Filtplot4(dfsv4,"Liu and West Filter SP500")
plotobs5<-Filtplot5(dfsv5,"Always Resampling SP500")
plotobs6<-Filtplot6(dfsv6,"No Resampling SP500")

ggarrange(plotobs1,plotobs2)
ggarrange(plotobs3,plotobs4)
ggarrange(plotobs5,plotobs6)

realdatasetSTOXX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
y2<-realdatasetSTOXX[,2]
timeframe<-c(1:length(y2))

sv_bpf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 6))[,1]
dfsv7<-data.frame(timeframe,y2,sv_bpf2)
dfsv8<-data.frame(timeframe,y2,sv_gpfopt2)
dfsv9<-data.frame(timeframe,y2,sv_apf2)
dfsv10<-data.frame(timeframe,y2,sv_lw2)
dfsv11<-data.frame(timeframe,y2,sv_bapf2)
dfsv12<-data.frame(timeframe,y2,sv_sis2)

Filtplot7<-function(dataframe,title){
  mean=sv_bpf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot8<-function(dataframe,title){
  mean=sv_gpfopt2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot9<-function(dataframe,title){
  mean=sv_apf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot10<-function(dataframe,title){
  mean=sv_lw2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot11<-function(dataframe,title){
  mean=sv_bapf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot12<-function(dataframe,title){
  mean=sv_sis2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotobs7<-Filtplot7(dfsv7,"Bootstrap Particle Filter STOXX50E")
plotobs8<-Filtplot8(dfsv8,"Opt Ker Guided Particle Filter STOXX50E")
plotobs9<-Filtplot9(dfsv9,"Auxiliary Particle Filter STOXX50E")
plotobs10<-Filtplot10(dfsv10,"Liu and West Filter STOXX50E")
plotobs11<-Filtplot11(dfsv11,"Always Resampling STOXX50E")
plotobs12<-Filtplot12(dfsv12,"No Resampling STOXX50E")

ggarrange(plotobs7,plotobs8)
ggarrange(plotobs9,plotobs10)
ggarrange(plotobs11,plotobs12)

realdatasetDJI<-read.xlsx("Dataset 17-21.xlsx", sheet = 3)
y3<-realdatasetDJI[,2]
timeframe<-c(1:length(y3))

sv_bpf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 6))[,1]
dfsv13<-data.frame(timeframe,y3,sv_bpf3)
dfsv14<-data.frame(timeframe,y3,sv_gpfopt3)
dfsv15<-data.frame(timeframe,y3,sv_apf3)
dfsv16<-data.frame(timeframe,y3,sv_lw3)
dfsv17<-data.frame(timeframe,y3,sv_bapf3)
dfsv18<-data.frame(timeframe,y3,sv_sis3)

Filtplot13<-function(dataframe,title){
  mean=sv_bpf3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot14<-function(dataframe,title){
  mean=sv_gpfopt3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot15<-function(dataframe,title){
  mean=sv_apf3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot16<-function(dataframe,title){
  mean=sv_lw3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot17<-function(dataframe,title){
  mean=sv_bapf3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot18<-function(dataframe,title){
  mean=sv_sis3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotobs13<-Filtplot13(dfsv13,"Bootstrap Particle Filter DJI")
plotobs14<-Filtplot14(dfsv14,"Opt Ker Guided Particle Filter DJI")
plotobs15<-Filtplot15(dfsv15,"Auxiliary Particle Filter DJI")
plotobs16<-Filtplot16(dfsv16,"Liu and West Filter DJI")
plotobs17<-Filtplot17(dfsv17,"Always Resampling DJI")
plotobs18<-Filtplot18(dfsv18,"No Resampling DJI")

ggarrange(plotobs13,plotobs14)
ggarrange(plotobs15,plotobs16)
ggarrange(plotobs17,plotobs18)

First, a general point is that the sequence of percentage log returns are different across the three indices. The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents. In particular, as previously noticed, there are extensive similarities between the DOW JONES Index and the S&P500, which can be visualized when looking at the percentage of log returns in the corresponding plots, since both are describing the US Economy. However, the STOXX50E differs quite extensively, especially in the different sensitivity to the Covid-19 Crisis and its consequences,  representing the European Economy. Aside from this note, the considerations expressed below on the sequences "filtered volatilities" for the various techniques hold almost uniformly for all the three indices. Therefore, in the next paragraph we will be referring to the set of plots pertaining to one specific index (which could be any of the three).\

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Riferimento alla sezione di convergenza?.} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles. Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below. Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling algorithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms. Finally, note that the Auxiliary Particle Filter employed here does not coincide with that of the section with synthetic data, which employed a different kernel implied by optimality results.}

```{r include=FALSE}
Errorcomp1<-matrix(NA,ncol=6,nrow=10)
colnames(Errorcomp1)<-c("N","BPF", "GPFOPT", "APF", "BAPF","SIS")
rownames(Errorcomp1)<-c("RMSE","RMSE","RMSE","RMSE","RMSE","MAE","MAE","MAE","MAE","MAE")
Errorcomp1[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
Errorcomp1[,2]<-c(0.3017986,0.11645976,0.02906876,0.01511962,0.010462488,0.2342816,0.08340162,0.02145381,0.01110398,0.008013466)
Errorcomp1[,3]<-c(0.3367199,0.09173166,0.03356642,0.01461339,0.009572002,0.2539500,0.06764348,0.02447722,0.01026689,0.007198043)
Errorcomp1[,4]<-c(0.4822315,0.1647978,0.05644653,0.02657352,0.02006171,0.3648792,0.1159503,0.03722911,0.01879724,0.01420517)
Errorcomp1[,5]<-c(0.4561433,0.1740267,0.04786161,0.02741662,0.02085289,0.3643665,0.1352035,0.03833600,0.01939683,0.01481767)
Errorcomp1[,6]<-c(1.566291,1.770554,1.1788696,0.9680993,0.9301619,1.256867,1.396812,0.8683638,0.7212099,0.7182857)

@@

Using the APF as the benchmark does not deliver a different picture than the one we previously had. Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at each threshold (though very slightly so for the higher ones). Its performance ameliorates dramatically at N=1000, while it still shows some sizable improvements even when passing from N=5000 to N=10000, almost halving its errors. Moreover, note that the mere fact that at N=10000 the Guided and the Bootstrap particle filters approximate better the APF benchmark suggests that the APF still has some margins for relative improvement even at high levels of N, pointing at higher computational costs than the competing filters: trivially, since at N=50000 its RMS and MA difference with respect to this benchmark must be zero, the APF overtakes the Guided and the Bootstrap PFs at N$>>10000$.\

```{r include=FALSE}
Errorcomp3<-matrix(NA,ncol=2,nrow=2)
colnames(Errorcomp3)<-c("RMSE", "MAE")
rownames(Errorcomp3)<-c("N=50000", "N=100000")
Errorcomp3[1,]<-c(0.015181843,0.008679359)
Errorcomp3[2,]<-c(0.013735482,0.006467285) #updated
```

@@

@@
